Positional surprisal (1-3grams) in CHILDES corpora

### 2019-05-30 surprisal/bnc&switchboard.Rmd
cleaned up surprisals folder, added childes and switchboard surprisal feathers

sw <- switchboard %>%
  left_join(switch_unigrams) %>%
  mutate(s = -log(p)) %>%
  ungroup() %>%
  select(length, word_order, s) %>%
  nest(s)

swc <- sw %>% filter(length %in% c(10, 20, 30))
switch_surprisals <- map_dfr(1:nrow(swc), 
        ~bind_cols(select(slice(swc, .x), length, word_order), 
                               tidyboot_mean(data = unnest(slice(swc, .x)), 
                                             column = s, nboot = 1000)))

# switch_surprisals <- switchboard %>%
#   left_join(switch_unigrams) %>%
#   mutate(s = -log(p)) %>%
#   ungroup() %>%
#   select(length, word_order, s) %>%
#   filter(length %in% c(10, 20, 30)) %>%
#   group_by(length, word_order) %>% 
#   tidyboot_mean(s)

switch_surprisals %>%
  ggplot(aes(x = word_order, y = empirical_stat, #observed mean surprisal
             ymin = ci_lower, ymax = ci_upper)) + 
  facet_wrap(~ length) + 
  ylab("Surprisal") + 
  xlab("Word Position in Switchboard") + 
  geom_pointrange(size = .50) + 
  geom_line(size = .50)
```

```{r switchboard bigrams}
switch_bigrams <- switchboard %>%
  group_by(utterance_id) %>%
  mutate(lag_word = lag(word)) %>%
  group_by(lag_word, word) %>%
  count() %>%
  filter(!is.na(lag_word)) %>%
  ungroup() %>%
  mutate(joint_p = n / sum(n)) %>%
  select(-n) %>%
  left_join(switch_unigrams, by = c("lag_word" = "word")) %>%
  mutate(cond_p = joint_p / p) %>%
  select(-n, -p)

@@

switch_bigram_surprisals %>%
  filter(length %in% c(5, 7, 9, 11, 13, 15)) %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)
```

```{r switchboard trigrams}
switch_trigrams <- switchboard %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  group_by(lag_word2, lag_word1, word) %>%
  count() %>%
  filter(!is.na(lag_word1), !is.na(lag_word2)) %>%
  ungroup() %>%
  mutate(tri_joint_p = n / sum(n)) %>%
  select(-n) %>%
  left_join(switch_bigrams, by = c("lag_word2" = "lag_word", "lag_word1" = "word")) %>%
  mutate(tri_cond_p = tri_joint_p / joint_p) %>%
  select(-joint_p, -tri_joint_p, -cond_p) # %>%

trigram_surprisals_prep <- switchboard %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  left_join(switch_trigrams) %>% 
  left_join(select(switch_bigrams, -joint_p), by = c("lag_word1" = "lag_word", "word" = "word")) %>% 
  left_join(switch_unigrams)

@@

switch_trigram_surprisals %>%
  filter(length %in% c(5, 7, 9, 11, 13, 15)) %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)
```

@@

bnc_surprisals %>%
  ggplot(aes(x = word_order, y = empirical_stat, #observed mean surprisal
             ymin = ci_lower, ymax = ci_upper)) + 
  facet_wrap(~ length) + 
  ylab("Surprisal") + 
  xlab("Word Position in BNC") + 
  geom_pointrange(size = .50) + 
  geom_line(size = .50)

# bnc_surprisals %>%
#   ggplot(aes(x = word_order, y = s)) + 
#   facet_wrap(~ length) + 
#   geom_point() + 
#   geom_smooth(se = F)
```

@@

bnc_bigram_surprisals <- bigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word), -log(p), -log(cond_p))) %>%
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)

bnc_bigram_surprisals %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)
```

>>>>

# 
# sw <- switchboard %>%
#   left_join(switch_unigrams) %>%
#   mutate(s = -log(p)) %>%
#   ungroup() %>%
#   select(length, word_order, s) %>%
#   nest(s)
# 
# swc <- sw %>% filter(length %in% c(10, 20, 30))
# switch_surprisals <- map_dfr(1:nrow(swc), 
#         ~bind_cols(select(slice(swc, .x), length, word_order), 
#                                tidyboot_mean(data = unnest(slice(swc, .x)), 
#                                              column = s, nboot = 1000)))

```

```{r switchboard bigrams}
switch_bigrams <- switchboard %>%
  group_by(utterance_id) %>%
  mutate(lag_word = lag(word)) %>%
  group_by(lag_word, word) %>%
  count() %>%
  filter(!is.na(lag_word)) %>%
  ungroup() %>%
  left_join(switch_unigrams, by = c("lag_word" = "word")) %>%
  mutate(cond_p = n.x / n.y) %>%
  select(-n.y, -p) %>%
  rename(n = n.x)

@@

```

```{r switchboard trigrams}
switch_trigrams <- switchboard %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  group_by(lag_word2, lag_word1, word) %>%
  count() %>%
  filter(!is.na(lag_word1), !is.na(lag_word2)) %>%
  ungroup() %>%
  left_join(switch_bigrams, by = c("lag_word2" = "lag_word", "lag_word1" = "word")) %>%
  mutate(tri_cond_p = n.x / n.y) %>%
  select(-n.x, -n.y, -cond_p)

trigram_surprisals_prep <- switchboard %>%
  group_by(utterance_id) %>%
  mutate(lag_word1 = lag(word)) %>%
  mutate(lag_word2 = lag(lag_word1)) %>%
  left_join(switch_trigrams) %>% 
  left_join(switch_bigrams, by = c("lag_word1" = "lag_word", "word" = "word")) %>% 
  left_join(switch_unigrams, by = c("word" = "word"))

@@

```

@@

```

@@

bnc_bigram_surprisals <- bigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word), -log(p), -log(cond_p))) %>%
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)
```

### 2019-05-30 surprisal/childes_surprisal.Rmd
cleaned up surprisals folder, added childes and switchboard surprisal feathers

Get utterances
```{r get_utterances}
prov_utterances <- get_utterances(corpus = "Providence") #stem or gloss
zhou_utterances <- get_utterances(corpus = "ZhouDinner") #stem for pinyin
prov_utterances <- get_utterances(corpus = "Okayama")
```

```{r old unigrams}
child_utterances <- prov_utterances %>%
  filter(speaker_role == "Target_Child") %>% 
  mutate(length = str_count(gloss, pattern = "[ +]+") + 1) %>%
  mutate(utterance_id = 1:n()) %>%
  unnest_tokens(word, gloss, token = stringr::str_split, pattern = "[ +]+") %>%
  group_by(utterance_id) %>%
  mutate(word_order = 1:n())

@@

surprise <- bind_rows(child_surprisals %>%
  mutate(speaker_role = "Child") %>%
  bind_rows(mutate(adult_surprisals, speaker_role = "Adult")) %>%
  mutate(Language = "Japanese"))

@@

```{r redoing bigrams}
adult_bigrams_prep <- adult_utterances %>% 
  group_by(utterance_id) %>%
  mutate(lag_word = lag(word)) %>%
  group_by(lag_word, word) %>%
  count() %>%
  filter(!is.na(lag_word)) %>%
  ungroup() %>%
  left_join(adult_unigrams, by = c("lag_word" = "word")) %>%
  mutate(cond_p = n.x / n.y) %>%
  select(lag_word, word, cond_p)

adult_bigrams <- adult_utterances %>%
  group_by(utterance_id) %>%
  mutate(lag_word = lag(word)) %>%
  ungroup() %>%
  left_join(adult_unigrams) %>%
  left_join(adult_bigrams_prep) %>%
  mutate(s = ifelse(is.na(lag_word), -log(p), -log(cond_p))) %>%
  group_by(length, word_order) %>%
  summarise(mean_s = mean(s))

label_data <- data_frame(length = 7, Language = "Japanese", word_order = 2,
                         speaker_role = c("Adult", "Child"), s = c(2,4))

```{r unigrams}
get_unigrams <- function(utterances, speaker_role, stem_or_gloss) {

  if (tolower(stem_or_gloss) == "stem") {
    utts <- utts %>%
      mutate(length = str_count(stem, pattern = "[ +]+") + 1) %>%
      mutate(utterance_id = 1:n()) %>%
      unnest_tokens(word, stem, token = stringr::str_split, pattern = "[ +]+") %>%
      group_by(utterance_id) %>%
      mutate(word_order = 1:n())

  } else {
    utts <- utts %>%
      mutate(length = str_count(gloss, pattern = "[ +]+") + 1) %>%
      mutate(utterance_id = 1:n()) %>%
      unnest_tokens(word, gloss, token = stringr::str_split, pattern = "[ +]+") %>%
      group_by(utterance_id) %>%
      mutate(word_order = 1:n())
  }

shiro_sups <- get_unigrams(shiro_utterances, "not child", "gloss") %>% 
  mutate(speaker_role = "adult") %>%
  bind_rows(mutate(get_unigrams(shiro_utterances, "child", "gloss"), speaker_role = "child")) %>%
  mutate(language = "Spanish")

zhou_sups <- get_unigrams(zhou_utterances, "not child", "stem") %>% 
  mutate(speaker_role = "adult") %>%
  bind_rows(mutate(get_unigrams(zhou_utterances, "child", "stem"), speaker_role = "child")) %>%
  mutate(language = "Mandarin")

all_sups <- prov_sups %>%
  bind_rows(shiro_sups, zhou_sups)

label_data <- data_frame(length = 9, Language = "Spanish", word_order = 3,
                         speaker_role = c("Adult", "Child"), s = c(1.25,3.5))

all_trigrams %>% 
  filter(length %in% c(5, 7, 9)) %>%
  ggplot(aes(x = word_order, y = s,
                        color = speaker_role,
                        label = speaker_role)) +
    facet_grid(Language ~ length, scales = "free") + 
    xlab("Word position") +
    ylab("Mean surprisal") +
    geom_line(size = 1) +
    geom_point() + 
    theme_classic(base_size = 10) + 
    scale_color_ptol() +
    theme(legend.position = "nones") + 
    geom_text(data = label_data)

@@

bigram_surprisals_nonnull <- bigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word), -log(p), -log(cond_p))) %>%
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)

all_bigrams <- all_bigrams %>% 
  bind_rows(mutate(bigram_surprisals_nonnull, speaker_role = "Child", Language = "Spanish"))

bigram_surprisals_nonnull %>%
  filter(length %in% c(5, 7, 9, 11, 13, 15)) %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)
```

@@

trigram_surprisals_nonnull <- trigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word2), # check if it's not the first word
                    ifelse(is.na(lag_word1), # check if not the second word
                           -log(p), # trigram if third or beyond
                           -log(cond_p)), # bigram if second word
                    -log(tri_cond_p))) %>% # unigram if first word
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)

trigram_surprisals_nonnull %>%
  filter(length %in% c(5, 7, 9, 11, 13, 15)) %>%
  ggplot(aes(x = word_order, y = s)) + 
  facet_wrap(~ length) + 
  geom_point() + 
  geom_smooth(se = F)

>>>>

Get utterances
```{r get_utterances}
prov_utterances <- get_utterances(corpus = "Providence") #stem or gloss
zhou_utterances <- get_utterances(corpus = "ZhouDinner") #stem for pinyin
okayama_utterances <- get_utterances(corpus = "Okayama")
```

```{r unigrams}
# column name must be a string, NOT just the name itself
get_unigrams <- function(utterances, speaker_role, name) {

  utts <- utts %>%
    rename(text = name) %>%
    mutate(length = str_count(text, pattern = "[ +]+") + 1) %>%
    mutate(utterance_id = 1:n()) %>%
    unnest_tokens(word, text, token = stringr::str_split, pattern = "[ +]+") %>%
    group_by(utterance_id) %>%
    mutate(word_order = 1:n())

Sample code for plotting your surprisal results.

```{r plotting}
label_data <- data_frame(length = 7, Language = "Japanese", word_order = 2,
                         speaker_role = c("Adult", "Child"), s = c(2,4))

Down here is the old code. You can run it, but you have to change the corpus name and whether you're using stem or gloss by hand.

```{r hard-coded language unigrams}
child_utterances <- prov_utterances %>%
  filter(speaker_role == "Target_Child") %>% 
  mutate(length = str_count(gloss, pattern = "[ +]+") + 1) %>%
  mutate(utterance_id = 1:n()) %>%
  unnest_tokens(word, gloss, token = stringr::str_split, pattern = "[ +]+") %>%
  group_by(utterance_id) %>%
  mutate(word_order = 1:n())

@@

@@

```{r new bigrams}
adult_bigrams_prep <- adult_utterances %>% 
  group_by(utterance_id) %>%
  mutate(lag_word = lag(word)) %>%
  group_by(lag_word, word) %>%
  count() %>%
  filter(!is.na(lag_word)) %>%
  ungroup() %>%
  left_join(adult_unigrams, by = c("lag_word" = "word")) %>%
  mutate(cond_p = n.x / n.y) %>%
  select(lag_word, word, cond_p)

adult_bigrams <- adult_utterances %>%
  group_by(utterance_id) %>%
  mutate(lag_word = lag(word)) %>%
  ungroup() %>%
  left_join(adult_unigrams) %>%
  left_join(adult_bigrams_prep) %>%
  mutate(s = ifelse(is.na(lag_word), -log(p), -log(cond_p))) %>%
  group_by(length, word_order) %>%
  summarise(mean_s = mean(s)) 
```

@@

bigram_surprisals <- bigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word), -log(p), -log(cond_p))) %>%
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)

@@

trigram_surprisals <- trigram_surprisals_prep %>%
  mutate(s = ifelse(is.na(lag_word2), # check if it's not the first word
                    ifelse(is.na(lag_word1), # check if not the second word
                           -log(p), # trigram if third or beyond
                           -log(cond_p)), # bigram if second word
                    -log(tri_cond_p))) %>% # unigram if first word
  group_by(length, word_order) %>%
  summarise(s = mean(s)) %>% 
  filter(word_order <= length)
```
